---
title: "Week 9 Lab Code"
author: "Section 019"
date: "2025-03-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Problem 1
### *qt()* Function
Use *qt()* function to get the corresponding $t$-value of $P(T \leq t) = p$ given the probability $p$ for a **t distribution** with the specified degree of freedom ($df$).  
Similar to *qnorm()* function, *qt()* function will return a $t$-value (percentile or quantile, q stands for quantile) corresponding to a given probability.  
Format: *qt(p, df = df)* (return $t$ such that $P(T \leq t) = p$ - left side probability)  
**Arguments**:  
*df*: degree of freedom, need to be explicitly specified in the command (because *df* is a compulsory parameter for t-distribution, and there is no default value of *df* in R). We usually use $df = n - 1$ when constructing confidence interval for $\mu$.  
*lower.tail*: extend *qt()* function for right side ($>$ or $\geq$) probability.  

What values of t-multiplier $t_{\alpha / 2}$ is used to construct:  

## a
a $92\%$ confidence interval to estimate $\mu$ if the sample size is $9$. (Assume that random sample assumption and normal population assumptions are met).  

### Answer
```{r, echo = FALSE}
curve(dt(x, df = 8),
      from = -3, to = 3,
      col = "black",
      xlab = "",
      ylab = "",
      main = "92% Confidence Interval, Sample Size = 9",
      lwd = 2)

x_high = seq(-2.004152, 2.004152, length.out = 200)
y_high = dt(x_high, df = 8)
polygon(x = c(-2.004152, x_high, 2.004152),
        y = c(0, y_high, 0),
        col = rgb(1, 0, 0, alpha = 0.4),
        border = NA)
```

Here, we want to find the t-multiplier $t$ such that $P(-t < T < t) = 0.92$, i.e., $P(T < -t) + P(T > t) = 1 - 0.92 = 0.08$. Since t distribution is symmetric, $P(T < -t) = P(T > t) = \frac{0.08}{2} = 0.04$.  
Since the sample size is $9$, the degree of freedom (argument *df*) of the t-distribution is $9 - 1 = 8$.  

We want to find the quantile $t$ that $P(T \leq t) = 1 - \frac{0.08}{2} = 0.96$.  
```{r}
qt(1 - 0.08 / 2, df = 8)
```

Similarly, $P(T \leq t) = 0.96$ is the same as $P(T > t) = 1 - 0.96 = 0.04$ (right tail probability, using the argument *lower.tail = FALSE*).  
```{r}
qt(0.08 / 2, df = 8, lower.tail = F)
```

Use the negative t-multiplier ($P(T < t) = \frac{0.08}{2} = 0.04$), but we usually use the positive t-multiplier to construct the confidence interval.  
```{r}
qt(0.08 / 2, df = 8)
```

## b
a $98\%$ confidence interval to estimate $\mu$ if the sample size is $1982$. (Assume that random sample assumption is met.)  

### Answer
Since the t-procedure is robust to non-normality if the sample is randomly drawn (because we do not know what the population distribution is), we can still use the t-distribution to constuct the confidence interval.   
```{r, echo = FALSE}
curve(dt(x, df = 1981),
      from = -3, to = 3,
      col = "black",
      xlab = "",
      ylab = "",
      main = "98% Confidence Interval, Sample Size = 1982",
      lwd = 2)

x_high = seq(-2.328232, 2.328232, length.out = 200)
y_high = dt(x_high, df = 1981)
polygon(x = c(-2.328232, x_high, 2.328232),
        y = c(0, y_high, 0),
        col = rgb(1, 0, 0, alpha = 0.4),
        border = NA)
```

Here, we want to find the t-multiplier $t$ such that $P(-t < T < t) = 0.98$, i.e., $P(T < -t) + P(T > t) = 1 - 0.98 = 0.02$. Since t distribution is symmetric, $P(T < -t) = P(T > t) = \frac{0.02}{2} = 0.01$.  
Since the sample size is $1982$, the degree of freedom (argument *df*) of the t-distribution is $1982 - 1 = 1981$.  

We want to find the quantile $t$ that $P(T \leq t) = 1 - \frac{0.02}{2} = 0.99$.  
```{r}
qt(1 - 0.02 / 2, df = 1981)
```

Similarly, $P(T \leq t) = 0.99$ is the same as $P(T > t) = 1 - 0.99 = 0.01$ (right tail probability, using the argument *lower.tail = FALSE*).  
```{r}
qt(0.02 / 2, df = 1981, lower.tail = F)
```

\newpage

# Problem 2
In this problem, we will use the class survey data set. You will need to import data set again within the scope of R markdown. For the sake of this problem, assume that the variable college GPA represents a random sample of 289 college students. We will calculate a $95\%$ confidence interval manually using $\bar{x}\pm t_{\alpha/2} \frac{s}{\sqrt{n}}$, then use `t.test()` command.  

Use the following command to import the survey data set.  
```{r}
survey = read.csv("http://users.stat.umn.edu/~parky/survey_S25.csv")
```

## a
Use `names( )` to see the names of all variables. What is the exact name of the variable that represents college GPA?  

### Answer
```{r}
names(survey)
```

The variable *c.gpa* is the variable that represents college GPA.  

## b
Find sample size ($n$), sample mean ($\bar{x}$, point estimate of population mean), sample standard deviation ($s$), standard error ($SE = \frac{s}{\sqrt{n}}$), t-multiplier ($t_{\alpha/2}$), margin of error ($t_{\alpha/2} \frac{s}{\sqrt{n}}$).  

### Answer
```{r}
#rename (since we want to use a shorter term for survey$c.gpa)
#create a new variable x to store survey$c.gpa
x = survey$c.gpa
```

**sample size**: number of observations in our sample.  
Use *length()* function to get the length of a vector (or a variable).  
Format: *length(vector_name)*  
```{r}
(n = length(x)) #R will print the outputs if we put () around a command
```

**sample mean** ($\bar{x}$, point estimate of population mean).  
```{r}
(xbar = mean(x))
```

**sample standard error**: $s$ (used to estimate the population standard deviation $\sigma$), measures how much a typical student's GPA varies from the sample mean from this sample of $289$ students.  
```{r}
(s = sd(x))
```

**standard error**: estimates how much typical **sample mean** (with sample size $289$) vary from the population mean when we repeatedly draw random samples. or average distance between a sample mean from $n = 289$ and the center of the sampling distribution of sample mean ($mu_{\bar{x}}$, which is the same as population mean $\mu$).  
$se = \frac{s}{\sqrt{n}}$, where $s$ is the **sample standard deviation**, $n$ is the **sample size**.  
```{r}
(se = s / sqrt(n))
```

**error probability** ($\alpha$): in the long run, the proportion of the times this method produces interval that doesn't contain the true parameter.  
```{r}
(alpha = 0.04)
```

**t-multiplier**: how many standard errors to add to the point estimate, computed by the *qt()* command.  
General command for any error probability ($\alpha$) and sample size ($n$): `t = qt(1 - alpha / 2, df = n - 1)`  
```{r}
(t = qt(1 - alpha / 2, df = n - 1))
```

or plug the numbers in:  
```{r}
(t = qt(1 - 0.04 / 2, df = 289 - 1))
```

**margin of error**: estimates how much most ($96\%$ in this example - confidence level) of sample means fall within from the true population mean - has the same unit as $x$.    
$\text{moe} = t_{\alpha / 2} \times SE$  
```{r}
(moe = t * se)
```

## c
Calculate a $96%$ confidence interval.  
Recall a $1 - \alpha$ confidence interval for $\mu$ is:  
$$\bar{x}\pm t_{\alpha/2}\frac{s}{\sqrt{n}}$$  

### Answer
Here $1 - \alpha = 96\%$, $\alpha = 0.04$.  
```{r}
#essential computation
xbar = mean(x) #sample mean
s = sd(x) #sample standard deviation
n = length(x) #sample size
alpha = 0.04 #error probability
t = qt(1 - alpha / 2, df = n - 1) #t-multiplier

#vetorized computation: c(-1, 1) * t * se adds and subtracts margin of error
xbar + c(-1, 1) * t * (s / sqrt(n))
```

$\bar{x} + \text{c(-1, 1)} \times t \times \frac{s}{\sqrt(n)}$ will create a vector $\left(\bar{x} + (-1) \times t \times \frac{s}{\sqrt(n)}, \bar{x} + 1 \times t \times \frac{s}{\sqrt(n)}\right) = \left(\bar{x} - t \times \frac{s}{\sqrt(n)}, \bar{x} + t \times \frac{s}{\sqrt(n)}\right)$  

The $96\%$ confidence interval is $[3.124419, 3.367159]$.  

\newpage

## d
Confidence interval for mean has two assumptions.  

- Random sample: We assume that the survey represents a random sample of college students. Generally, to check the random sample assumption, we need to know how the observations were selected.  

- Normality assumption: To determine whether the population distribution is normal or not, we use data distribution.  If data distribution doesn't deviate much from the normal distribution, normality assumption is met.  

Construct a Q-Q plot to determine whether the normality assumption is met. If it is not met, is our result still reliable? (Does this method work $96\%$ of the time?)  

### Answer
Use *qqnorm()* function to draw the Q-Q plot.  
Use *qqline()* function to draw the Q-Q line (used to check normality).  
Format: *qqnorm(vector_name)*, *qqline(vector_name)*; run *qqnorm()* function first  
```{r}
qqnorm(x)
qqline(x)
```

The overall distribution is left-skewed and not normal since most of the observaions are below the Q-Q line. Since there are a lot of $0$'s and $4$'s, the distribution might be bimodal.  
However, we can still be confident in our result because when sample size $n = 289$ is large and there is no extreme outliers (robustness of t-procedure; sampling distribution is approximately normal by CLT when $n = 289 > 30$)  

\newpage

## e
Use `t.test()` command to construct a 96\% confidence interval. Report the confidence interval and interpret the result.  

### *t.test()* Function
Use *t.test()* function to construct confidence interval (what we will do here) and perform hypothesis test (in the next few weeks).  
Construct Confidence Interval: *t.test(variable_name, conf.level = 1 - alpha, alternative = "two.sided")*  
Arguments:  
*conf.level*: confidence level, $1 - \alpha$ for the $1 - \alpha$ confidence interval  
*alternative*: confidence interval is **two sided**  
**Attention**:  
Always set the argument *alternative = "two.sided"* when constructing confidence interval.  
Put the argument *two.sided* in a pair of quotation marks because it is a textal argument (like the title of graphs).  

### Answer
Here $1 - \alpha = 96\%$, $\alpha = 0.04$.  
```{r}
t.test(x, conf.level = 1 - 0.04, alternative = "two.sided")
```

**Interpretation**:  
If we **repeatedly** select a random sample and construct a confidence interval, then **in the long run**, $96 \%$ of those intervals would **contain** the **true mean** of the variable college GPA.  
  
OR  
  
**In the long run**, $96 \%$ of the $96 \%$ confidence interval will **contain** the **true mean** of the variable college GPA.  
  
OR  
  
We are $96 \%$ confident that the interval $[3.124419, 3.367159]$ **contains** the **true population mean** of the variable college GPA.  

## f
If we increases the confidence level from $96\%$ to $99\%$, holding everything else fixed, which of the following is false?  

- (i) Margin of error increases.    
- (ii) Width of the confidence interval increases.  
- (iii) Standard error increases.  
- (iv) Point estimate does not change.  

### Answer
(i) is true. Margin of error $t_{\alpha/2} \frac{s}{\sqrt{n}}$ increases when confidence level increases because t-multiplier increases.  
```{r}
# for 96% confidence level
alpha = 0.04
n = 289
qt(1 - alpha / 2, df = n - 1)

# for 99% confidence level
alpha = 0.01
qt(1 - alpha / 2, df = n - 1)
```

t-multipler increase from $2.063092$ to $2.593008$.  
  
(ii) is true. Width of the confidence interval is determined by margin of error ($\text{width} = \text{upper bound} - \text{lower bound} = \bar{x} + moe - (\bar{x} - moe) = 2 \times moe$. When margin of error increases, confidence interval becomes wider.  
  
(iii) is false. Standard error $\frac{s}{\sqrt{n}}$ does not change when confidence level changes (no $\alpha$ in the formula, margin of error - $moe = t_{\alpha / 2} \times SE$ will change).     
  
(iv) is correct. Point estimate of population mean is an observed sample mean $\bar{x}$, which does not change when confidence level increases.  