---
title: "Week 1 Lab Code"
author: "Section 019"
date: "2025-01-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Use R Console as a calculator
Hit \textit{Enter} (or \textit{Return}) to run the code in R Console.  
Run only one line of command in R Console each time hitting Enter or Return.  
```{r}
9 + 7
```

```{r}
sqrt(16)
```
\textit{sqrt()} function is used to compute the square root of a number (in decimals if the answer is not an integer).
Format: \textit{sqrt(n)} $= \sqrt{n}$  

```{r}
abs(-3)
```
\textit{abs()} function is used to compute the absolute value of a number.  
Format: \textit{abs(n)} $= |n|$

```{r}
3 ^ 2
3 ** 2
```
Two ways for the exponentiation.  
$m \textasciicircum n = m ** n = m ^ n$

```{r}
#3 % 2
#This is an invalid command in R
```
Error: unexpected input in "3 % 2"
No % operation in R.  
We usually use / (divided by) for division method.  

```{r}
3 / 2
```

If we want to use the modular arithmetic (get the remainders), we need to use %%.
```{r}
3 %% 2
```
$3 = 2 * 1 + 1$

```{r}
3e1
3e-1
```
\textit{e}: $* \, 10$ to the power of  
$men = m * 10 ^ n$  
e.g., $3e1 = 3 * 10 ^ 1 = 3 * 10 = 30$, $3e-1 = 3 * 10 ^ {-1} = 3 * 0.1 = 0.3$  

\newpage

# Create Objects in R (in Code Editor)
```{r}
x <- 1 #x = 1
y = 3 #y <- 3
xy <- x * y #xy = x * y
```

## Print the Values of Objects
```{r}
x
y
xy
```
Directly use the names of objects to print the values of these objects.  

## Assign Values to Objects
= and <- are used to assign values to objects.  
Assign the \textbf{value} on the \textbf{right} to an \textbf{object} with the \textbf{name} on the \textbf{left}.  
= and <- have \textbf{exactly} the same function.
Format: \textit{object name} = or <- \textit{object value}  

## Run R Codes
In \textbf{Code Editor}, hitting Enter (or Return) will move the cursor into a new line.  
In \textbf{R Console}, hitting Enter (or Return) will run one line of code.  

### Run the Codes in Code Editor (3 Methods)
1. Select all liens of codes, then click the Run button on the top right of the Code Editor window.  
2. Hit \textbf{Control + Enter} (or \textbf{Command + Return}) when have finished entering each line of code.  
3. Select all liens of codes, then hit \textbf{Control + Enter} (or \textbf{Command + Return}).  

\newpage

# Create Vectors and Making Plots
R is a \textbf{vector-based} programming language.  

## Add Comments in R
Use # to add comments, and R will not consider comments as codes or run them.  
Each time adding a comment, always start with #.  
One # only works in one line (the same line after #).  
```{r}
x <- 1 #add a comment after code
# add a comment line
# add another comment line
```

## Create Vectors
```{r}
# monthly avg high temp
# <- and = are equal when assigning values
tempt <- c(24, 28, 42, 56, 69, 79, 84, 81, 73, 58, 42, 29)
tempt = c(24, 28, 42, 56, 69, 79, 84, 81, 73, 58, 42, 29)

tempt #print the vector tempt
```
Function \textit{c()} is used to create a vector by combining many single objects.  
"c" stands for combine.  
Format: \textit{c($n_1, n_2, \cdots$)}  

\newpage

## Draw Plots
```{r}
#Tempt
#plot(Tempt)
```
Error: object 'Tempt' not found  
R is sensitive to the case of object names.  
Tempt and tempt are \textbf{different} vectors.  

```{r}
tempt #print the vector tempt
plot(tempt)
```

Function \textit{plot()} is used to draw a scatter plot.  
Format: \textit{plot(vector)}  

\newpage

If we change the values in the vector \textit{tempt}, the scatter plot created by the function \textit{plot()} will also change.  
```{r}
tempt = c(40, 28, 42, 56, 69, 79, 84, 81, 73, 58, 42, 50) #change the first and the last element of the vector
tempt #print the values of the new vector tempt
plot(tempt)
```

### Use Help Documents in R
```{r}
?plot
```

Format: ?\textit{function\_name}  
Get the help document in the \textbf{Files \& Plots} window.
